\textsc{\wobble: a data-driven method for precision radial velocities}

### 2018-11-29 paper/paper.tex
making some suggested changes

\begin{abstract}
% Context
Extreme-precision radial velocity (\EPRV) exoplanet surveys face considerable data analysis challenges in extracting maximally precise \RVs from spectra.
Chief among these, particularly for the upcoming generation of red-optimized spectrographs targeting M dwarfs, is the presence of telluric absorption features, which are not perfectly known.
Another major limitation on the achievable \RV precision is the need to adopt an imperfect stellar template against which to cross-correlate or otherwise match the observations.
In both cases, the precision-limiting reliance on external information can be sidestepped using the data itself.
% Aims
Here we propose a data-driven method to simultaneously extract precise \RVs and infer the underlying stellar and telluric spectra using a linear model (in the log of flux).
The model employs a convex objective and convex regularization to keep the optimization of the spectral components fast.
We implement this method in \wobble, an open-source \python package which uses \TF in its first non-neural-network application to astronomical data.
In this work, we demonstrate the performance of \wobble on archival \HARPS spectra.
% Results
We recover the canonical exoplanet 51 Pegasi b, detect the secular \RV evolution of the M dwarf \Mdwarf, and retrieve the \RM for the Hot Jupiter HD 189733b.
We also present extremely high-\SNR composite stellar spectra and detailed time-variable telluric spectra derived from these data.
\end{abstract}

@@

For the case of the telluric component, whose spectrum is allowed to vary with time, we assume this spectral variability is low-dimensional.
This assumption is physically motivated in the sense that that a relatively small number of molecular species contribute to the telluric absorption spectrum.
It is also needed in a practical sense, since every additional dimension over which the telluric spectrum can vary adds several thousand more free parameters to the model.

@@

Aside from this regularization to push model components to zero in the low-\SNR regime, the model makes no assumptions about the shape of the stellar or telluric spectra.
We solve for spectral templates for each component as a series of control points with no imposed correlations between them, meaning that the line spread function, covariances between lines arising from the same species, and other such physically expected correlations are not built into the model but must be learned in the process of optimizing.
In addition to keeping this model simple, linear, and largely convex
(\todo{DFM says: what does ``largely'' mean here?}), this means that no physical knowledge about the object being observed is needed to extract its \RVs.

For practical applications, the number of epochs $N$ must be large (\todo{MB
-- how large? 12? 50? 100? Probably good to give some sense}) and span a significant fraction of the observing season.
If the number of spectra is small or if all spectra are taken over a very small range of time (e.g. a single night), it will be nearly impossible to disentangle the components of the model, as this process relies on the apparent Doppler shifting of the star(s) and Earth's atmosphere relative to each other.
We will revisit this restriction in Section \ref{s:future}.

@@

The stellar spectrum contribution is:
\begin{equation}
y_{n}^\star = P(v_{obs, n}) \mu_{\star},
\end{equation}
where $\mu_{\star}$ is a spectral template with length $M'$ corresponding to
the length of an arbitrarily-chosen wavelength grid of control points, and
$P$ is a linear operator which applies a Doppler shift by velocity $v_n$ and
interpolates $\mu_\star$ into the observatory rest frame.
Because we work in log-wavelength space, the Doppler shift due to velocity $v$ is simply an additive term:
\begin{equation}
 \ln \lambda(v) = \ln \lambda_{0} + \frac{1}{2} \ln \left(\frac{1 - v/c}{1 + v/c}\right).
\end{equation}
The $P$ operator adds this term to the $M'$-length wavelength grid for the template spectrum before linearly interpolating the template to the $M$-length data space.
\todo{DFM says: feel free to cut this:} $P$ could be represented as a rectangular
$M \times M'$ sparse matrix, but the dot product can be implemented
efficiently using linear interpolation instead of instantiating the full matrix.

@@

After subtracting the \RVs as predicted by the equations of motion at each epoch, the residuals have low dispersion, as expected for a quiet star with no known planets.
The \acronym{RMS} scatter among \wobble\ \RVs is 2.0 \ms.
This compares favorably to the scatter of 2.5 \ms among \RVs produced by the standard closed-source \HARPS pipeline, although we note that many of the residuals have similar non-zero values in both independently produced RV estimates.
It is possible that these deviations are a real physical effect in the stellar spectrum, but we caution that approximately \ms-level errors are likely being introduced by the barycentric correction.
Currently we assume that the \BERV provided by the native \HARPS pipeline is correct, but this will not hold true if the sky coordinates entered by the observer deviate from the actual on-sky location of the target.
This is especially likely to be the case for high proper motion targets like \Mdwarf.
Indeed, we do note a strong peak at year-long periods in a Lomb-Scargle periodogram of the residual \RVs.
Thus the \RV scatter derived in this work is only an approximate upper limit on the true precision achievable by \wobble. \todo{(TO DO: recalculate \BERV ourselves)}

@@

One consequence of working in the low-\SNR regime is that the power to resolve very small telluric features is reduced.
To keep the template spectra from overfitting the noise, the regularization amplitudes on the templates must be raised by several orders of magnitude relative to what was optimal for the 51 Peg fit above, particularly in orders where no strong spectral features are available.
This has the side effect of flattening out weak features if no strong lines are present in a given order.
As a result, we are able to retrieve fewer telluric lines in the bluer orders (compare the top panels of Figures \ref{fig:51peg_spectrum} and \ref{fig:barnards_spectrum} \todo{-- To do: check if there's a better example order}).
In redder orders, where the \SNR is higher and more strong telluric lines are present, the features are retrieved (bottom panel of Figure \ref{fig:barnards_spectrum}).
However, telluric variability was not able to be resolved in the overwhelming majority of orders; in fact, the results presented here were obtained with the tellurics model set to $K=0$ (no variability included).
We therefore caution that difficult-to-resolve microtellurics and time-variable features may not be reliably disentangled from the stellar spectrum when \wobble is applied to lower signal observations \SNR $\lessapprox 50$ pixel$^{-1}$).

Despite the extreme noise at the bluer end of the wavelength range, where \Mdwarf is very faint, \wobble successfully retrieves a template spectrum that appears consistent with general expectations for a mid-M dwarf through most of the \HARPS wavelength range.
As a sanity check, we compare the \wobble stellar template model with a high-resolution \PHOENIX model at \Mdwarf's previously measured spectral parameters \citep[\teff = 3200 K, \logg = 5.0, and \mh = $-0.5$;][and references therein]{Husser2013, Artigau2018}.
Even in the bluest regions, the features inferred by \wobble compare well with the \PHOENIX predictions in broad strokes (Figure \ref{fig:barnards_model}).
\todo{(TO DO: ask Travis Barman about opacity binning affecting the detailed features)}
\todo{DFM says: when I look at figure 5, I feel like the two spectra really don't
look similar. I think this could use some more discussion about why we find
this figure satisfying.}

@@

Along with minimum requirements on the spectral data quality, \wobble has some limitations driven by the number of spectra available and the sampling of those spectra.
Some minimum quantity of spectra are necessarily simply because the spectral templates contribute many free parameters, deriving up the data points needed for inference.
The sampling of those spectra are also important because the stellar lines must undergo substantial \RV variations to disentangle them from the telluric features, which is only feasibly achieved by observing throughout the season to take advantage of the $\sim 30$ \kms \BERV shift.
This requirement could be partially mitigated by modeling multiple stars with a shared telluric spectral template, which we discuss in Section \ref{s:improvements}.
However, in general the \wobble method fundamentally requires a substantial number of high quality spectra to perform reliably.

@@

%Spectral variability due to LSF changing? Additional parameters to the P operators, this would go into the non-convex optimization part (ask Hogg to throw in some words earlier on about location vs linear parameters)
Another way a less well-calibrated instrument may violate \wobble's assumptions is in the non-negligible variability of all spectral lines due to changes in the instrumental line spread function (\LSF).
In principle, if the changes to the \LSF were well-understood they could be hard-coded into the P operators.
Realistically speaking, if \LSF variations are present they are probably not sufficiently well-understood to take this approach; instead, we might want to fit these variations as a data-driven model component.
This leads us into the territory of making more fundamental changes to the \wobble model.

@@

One important feature of \wobble is that it models all fluxes in the log-space.
When the model consists solely of multiplicative components, as is the case for a single star + telluric absorption, this choice makes the optimization convex at fixed \RV and computationally tractable \todo{(HOGG: edit plz)}.
If we chose instead to convert the $y$ data vectors to linear fluxes and use a model consisting only of additive components, the same equations and algorithms would apply.
Example applications of such a model might be a multiple star system or a faint star with substantial sky background present in the observed spectra.
If, however, the model needed to incorporate both additive and multiplicative
components, major structural changes would be necessary and much of the
optimization speed would be lost \todo{DFM says: this is not obvious to me.
We're not currently dependent on anything being convex. I don't think we
should say this without trying it...}.
This means that at present, additive components may only be easily modeled by either fixing or neglecting the multiplicative components.
\todo{(HOGG: say things about hacks - maybe in a footnote even?)}

One extra model component that is at present both simpler to implement and perhaps more important to the results than additive components is the spectral continuum.
We have assumed that our simple continuum normalization is good, but this is certainly not true at a detailed level.
To some extent, this wrongness is absorbed into the telluric spectrum, as both the tellurics and the continuum can be treated as observatory rest frame features.
However, since the continuum and tellurics vary through time in different ways, the continuum should in principle be its own independent model component.
This component will be highly degenerate with the tellurics, but a clever representation could circumvent this: for example, the continuum spectrum could be placed on an extremely sparse wavelength grid to ensure that very localized effects like absorption lines are not included.

@@

We have also assumed that the stellar spectrum is constant, which is almost certainly false.
A variety of physical effects cause changes in the spectrum including stellar oscillations, convection, and activity in the form of starspots or plages.
\wobble can model stellar variability trivially by making the stellar component of the model take a similar form to the tellurics, where a set of basis vectors and weights are inferred from the data to capture variability.
This \PCA-like approach may be quite effective \citep{Davis2017}.
It does, however, introduce potential degeneracies between the inferred stellar variability and the Doppler shift.
To avoid this degeneracy, we might instead represent the stellar spectrum with an explicit parameterization of the line profile and restrict variability to the line depths, widths, and skews; such a change would fall along the same lines as the global \LSF parameterization proposed in Section \ref{s:data-changes}.

%Some absorption lines are more sensitive to the effects of stellar activity than others (Dumusque). If these fall into a small number of categories, a PCA-like dimensionality reduction should do well. Address pros and cons of introducing Gaussian processes to parameterize line profiles. deal with this similarly to the LSF changing? also, maybe we need to regularize (or put priors on) the \RVs themselves and force them to sum of Keplerians for the purpose of breaking degeneracies with stellar variability.

@@

One choice that \wobble makes differently from other data-driven \RV codes is the way we represent the underlying spectral components.
We use an extremely simple representation: a model grid of wavelengths and corresponding fluxes that gets linearly interpolated by the P operator.
One possible improvement would be to go to a higher-order interpolation scheme.
Another approach would be a non-parameteric method like a Gaussian process (as in the \code{STARFISH} code from \citet{Czekala2017}).
A third approach would be to use a model that incorporates more physics: this could range from a model with fixed lines that may vary only in terms of common line profile parameters to a complex model including various atomic and molecular physics of the lines.
A sensible way to implement this within the existing \wobble framework might be to add a physically motivated mean spectral template and optimize $y_{\star, n}$ vectors of residuals away from this template.
In any case, the general approach of \wobble does not depend intrinsically on the spectral representation, and swapping out the current representation for something else is simple, provided that the representative function can be implemented in \TF.

@@

Finally, a change that is quite major but worth stating regardless is the possibility of fitting the data directly in the 2D domain.
Widely used methods of extracting 1D spectra almost certainly sacrifice some degree of spectral information \citep{Bolton2010}.
In the future, we may wish to move to fitting the data directly without extracting.
The equations for \wobble would change very little in this regime: essentially, the P operator would become a 2D interpolator.
However, implementation of this model would become much more complicated, and extra variable components like sky background and spectrograph dispersion functions would need to be introduced.
While this implementation is far beyond the scope of current work, we nevertheless emphasize that the \wobble model is a useful starting framework in which to work towards this goal.

\vspace{2cm}

>>>>

\begin{abstract}
% Context
Extreme-precision radial velocity (\EPRV) exoplanet surveys face considerable data analysis challenges in extracting maximally precise \RVs from spectra. 
Chief among these, particularly for the upcoming generation of red-optimized spectrographs targeting M dwarfs, is the presence of telluric absorption features which are not perfectly known. 
Another major limitation on the achievable \RV precision is the need to adopt an imperfect stellar template against which to cross-correlate or otherwise match the observations. 
In both cases, the precision-limiting reliance on external information can be sidestepped using the data itself. 
% Aims
Here we propose a data-driven method to simultaneously extract precise \RVs and infer the underlying stellar and telluric spectra using a linear model (in the log of flux). 
The model employs a convex objective and convex regularization to keep the optimization of the spectral components fast. 
We implement this method in \wobble, an open-source \python package which uses \TF in one of its first non-neural-network applications to astronomical data. 
In this work, we demonstrate the performance of \wobble on archival \HARPS spectra.
% Results
We recover the canonical exoplanet 51 Pegasi b, detect the secular \RV evolution of the M dwarf \Mdwarf, and retrieve the \RM for the Hot Jupiter HD 189733b. 
We also present extremely high-\SNR composite stellar spectra and detailed time-variable telluric spectra derived from these data. 
\end{abstract}

@@

For the case of the telluric component, whose spectrum is allowed to vary with time, we assume this spectral variability is low-dimensional. 
This assumption is physically motivated in the sense that a relatively small number of molecular species contribute to the telluric absorption spectrum. 
It is also needed in a practical sense, since every additional dimension over which the telluric spectrum can vary adds several thousand more free parameters to the model.

@@

Aside from this regularization to push model components to zero in the low-\SNR regime, the model makes no assumptions about the shape of the stellar or telluric spectra. 
We solve for spectral templates for each component as a series of control points with no imposed correlations between them, meaning that the line spread function, covariances between lines arising from the same species, and other such physically expected correlations are not built into the model but must be learned in the process of optimizing. 
In addition to keeping this model simple and linear, this means that no physical knowledge about the object being observed is needed to extract its \RVs.

For practical applications, the number of epochs $N$ must be large ($N \gg N_{components}$). 
This is a fundamental restriction rooted in the fact that every spectrum being modeled introduces a large number of free parameters to the model. 
In most cases, these epochs will also need to span a significant fraction of the observing season. 
In order to disentangle the stellar and telluric spectral components, the spectra must undergo Doppler shifts with respect to each other that are at least as large as a resolution element of the spectrograph. 
For stars which do not physically undergo large \RV shifts over very short timescales, this means that it is necessary to observe the star over a significant fraction of the year to take advantage of the changes in the Earth's projected motion. 
We will comment on ways to overcome this restriction in Section \ref{s:future}.

@@

The stellar spectrum contribution is: 
\begin{equation}
y_{n}^\star = P(v_{obs, n}) \mu_{\star},
\label{eqn:star}
\end{equation}
where $\mu_{\star}$ is a spectral template with length $M'$ corresponding to
the length of an arbitrarily-chosen wavelength grid of control points, and
$P$ is a linear operator which applies a Doppler shift by velocity $v_n$ and
interpolates $\mu_\star$ into the observatory rest frame.
Because we work in log-wavelength space, the Doppler shift due to velocity $v$ is simply an additive term:
\begin{equation}
 \ln \lambda(v) = \ln \lambda_{0} + \frac{1}{2} \ln \left(\frac{1 - v/c}{1 + v/c}\right).
\end{equation}
The $P$ operator adds this term to the $M'$-length wavelength grid for the template spectrum before linearly interpolating the template to the $M$-length data space.
$P$ could be represented as a rectangular $M \times M'$ sparse matrix, but the dot product can be implemented efficiently using linear interpolation instead of instantiating the full matrix.

@@

After subtracting the secular \RV trend, the residuals have low dispersion, as expected for a quiet star with no planetary signals above $K \sim 1.2~\ms$ \citep{Choi2013, Ribas2018}. 
The \acronym{RMS} scatter among \wobble\ \RVs is 2.0 \ms. 
This compares favorably to the scatter of 2.5 \ms among \RVs produced by the standard closed-source \HARPS pipeline, although we note that many of the residuals have similar non-zero values in both independently produced RV estimates. 
It is possible that these deviations are a real physical effect in the stellar spectrum, but we caution that approximately \ms-level errors are likely being introduced by the barycentric correction. 
Currently we assume that the \BERV provided by the native \HARPS pipeline is correct, but this will not hold true if the sky coordinates entered by the observer deviate from the actual on-sky location of the target. 
This is especially likely to be the case for high proper motion targets like \Mdwarf. 
Indeed, we do note a strong peak at year-long periods in a Lomb-Scargle periodogram of the residual \RVs. 
Thus the \RV scatter derived in this work is only an approximate upper limit on the true precision achievable by \wobble. \todo{(TO DO: recalculate \BERV ourselves)} 
%\todo{How does HARPS calculate the BERVs? If there is something like cirrus, where the apparent brightness of the star varies over the observation, it is plausible that the flux-weighted observation midpoint and the actual time midpoint are different enough to cause a m/s difference in the observations, too. Does HARPS have a way of accounting for this? (HIRES does estimate this, and they have shorter exposures on average).}

@@

One consequence of working in the low-\SNR regime is that the power to resolve very small telluric features is reduced. 
To keep the template spectra from overfitting the noise, the regularization amplitudes on the templates must be raised by several orders of magnitude relative to what was optimal for the 51 Peg fit above, particularly in orders where no strong spectral features are available. 
This has the side effect of flattening out weak features if no strong lines are present in a given order. 
As a result, we are able to retrieve fewer telluric lines in the bluer orders (compare the top panels of Figures \ref{fig:51peg_spectrum} and \ref{fig:barnards_spectrum} \todo{-- To do: check if there's a better example order}). 
In redder orders, where the \SNR is higher and more strong telluric lines are present, the features are retrieved (bottom panel of Figure \ref{fig:barnards_spectrum}). 
However, telluric variability was not able to be resolved in the overwhelming majority of orders; in fact, the results presented here were obtained with the tellurics model set to $K=0$ (no variability included). 
We therefore caution that difficult-to-resolve microtellurics and time-variable features may not be reliably disentangled from the stellar spectrum when \wobble is applied to lower signal observations (\SNR $\lessapprox 50$ pixel$^{-1}$).

Despite the extreme noise at the bluer end of the wavelength range, where \Mdwarf is very faint, \wobble successfully retrieves a template spectrum that appears consistent with general expectations for a mid-M dwarf through most of the \HARPS wavelength range. 
As a sanity check, we compare the \wobble stellar template model with a high-resolution \PHOENIX model at \Mdwarf's previously measured spectral parameters \citep[\teff = 3200 K, \logg = 5.0, and \mh = $-0.5$;][and references therein]{Husser2013, Artigau2018}. 
Even in the bluest regions, the placements and relative strengths of the large-amplitude features inferred by \wobble generally compare well with the \PHOENIX predictions (Figure \ref{fig:barnards_model}). 
The \wobble model for \Mdwarf diverges from the theoretical models in the smaller absorption lines and in the degree of broadening for strong lines.  
\todo{(TO DO: are these known model weaknesses? ask Travis Barman about opacity binning affecting the detailed features)}

@@

Along with minimum requirements on the spectral data quality, \wobble has some limitations driven by the number of spectra available and the sampling of those spectra. 
Some minimum quantity of spectra are necessarily simply because the spectral templates contribute many free parameters, driving up quantity of data needed for inference. 
The sampling of those spectra are also important because the stellar lines must undergo substantial \RV variations to disentangle them from the telluric features, which for most targets can only be achieved by observing throughout the year to take advantage of the $\sim 30$ \kms~\BERV shift. 
This requirement could be partially mitigated by modeling multiple stars with a shared telluric spectral template, which we discuss in Section \ref{s:improvements}. 
However, in general the \wobble method fundamentally requires a substantial number ($N \gtrsim 10$) of high quality spectra to perform reliably.

@@

%Spectral variability due to LSF changing? Additional parameters to the P operators, this would go into the non-convex optimization part (ask Hogg to throw in some words earlier on about location vs linear parameters)
Another way a less well-calibrated instrument may violate \wobble's assumptions is in the non-negligible variability of all spectral lines due to changes in the instrumental line spread function (\LSF). 
In principle, if the changes to the \LSF were well-understood they could be hard-coded into linear operators like the $P$ operator used in Equation \ref{eqn:star}, leaving the essential equations of \wobble largely unchanged. 
Realistically speaking, if \LSF variations are present they are probably not sufficiently well-understood to take this approach; instead, we might want to fit these variations as a data-driven model component. 
This leads us into the territory of making more fundamental changes to the \wobble model.

@@

One important feature of \wobble is that it models all fluxes in the log-space. 
When the model consists solely of multiplicative components, as is the case for a single star + telluric absorption, this choice makes the optimization convex at fixed \RV and computationally tractable \todo{(HOGG: edit plz)}. 
If we chose instead to convert the $y$ data vectors to linear fluxes and use a model consisting only of additive components, the same equations and algorithms would apply. 
Example applications of such a model might be a multiple star system or a faint star with substantial sky background present in the observed spectra. 
If, however, the model needed to incorporate both additive and multiplicative
components as free variables, larger structural changes would be needed in the code and the convexity of the template optimization would be lost. 
Further tests would be needed to determine whether this loss of convexity would slow down the optimization.

One extra model component that is at present both simpler to implement and perhaps more important to the results than generic additive components is the spectral continuum. 
We have assumed that our simple continuum normalization is good, but this is certainly not true at a detailed level. 
To some extent, this wrongness is absorbed into the telluric spectrum, as both the tellurics and the continuum can be treated as observatory rest frame features. 
However, since the continuum and tellurics vary through time in different ways, the continuum should in principle be its own independent model component. 
This component will be highly degenerate with the tellurics, but a clever representation could circumvent this: for example, the continuum spectrum could be placed on an extremely sparse wavelength grid to ensure that very localized effects like absorption lines are not included.

@@

We have also assumed that the stellar spectrum is constant, which is almost certainly false. 
A variety of physical effects cause changes in the spectrum including stellar oscillations, convection, and activity in the form of starspots or plages. 
\wobble can model stellar variability trivially by making the stellar component of the model take a similar form to the tellurics, where a set of basis vectors and weights are inferred from the data to capture variability. 
This \PCA-like approach may be quite effective, since stellar features with similar line formation physics likely co-vary in the spectrum \citep{Davis2017, Dumusque2018}. 
It does, however, introduce potential degeneracies between the inferred stellar variability and the Doppler shift. 
To avoid this degeneracy, we might instead represent the stellar spectrum with an explicit parameterization of the line profile and restrict variability to the line depths, widths, and skews; such a change would fall along the same lines as the global \LSF parameterization proposed in Section \ref{s:data-changes}.

@@

One choice that \wobble makes differently from other data-driven \RV codes is the way we represent the underlying spectral components. 
We use an extremely simple representation: a model grid of wavelengths and corresponding fluxes that gets linearly interpolated by the $P$ operator. 
One possible improvement would be to go to a higher-order interpolation scheme. 
Another approach would be a non-parameteric method like a Gaussian process \citep[as in][]{Czekala2017}. 
A third approach would be to use a model that incorporates more physics: this could range from a model with fixed lines that may vary only in terms of common line profile parameters to a complex model including various atomic and molecular physics of the lines. 
A sensible way to implement this within the existing \wobble framework might be to add a physically motivated mean spectral template and optimize $y_{\star, n}$ vectors of residuals away from this template. 
In any case, the general approach of \wobble does not depend intrinsically on the spectral representation, and swapping out the current representation for something else is simple, provided that the representative function can be implemented in \TF.

@@

Finally, a change that is quite major but worth stating regardless is the possibility of fitting the data directly in the 2D domain. 
Widely used methods of extracting 1D spectra almost certainly sacrifice some degree of spectral information \citep{Bolton2010}.
In the future, we may wish to move to fitting the data directly without extracting. 
The equations for \wobble would change very little in this regime: essentially, the $P$ operator in Equation \ref{eqn:star} would become a 2D interpolator. 
However, implementation of this model would become much more complicated, and extra variable components like sky background and spectrograph dispersion functions would need to be introduced. 
While this implementation is far beyond the scope of current work, we nevertheless emphasize that the \wobble model is a useful starting framework in which to work towards this goal.
